---
title: "R Code & Resources"
date: 'Last updated: `r format(Sys.time(), "%d %B %Y")`'
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

(update text) The following page provides examples for various coding and quantitative techniques I've used over the years that may be useful to others. This page is a *work in progress*. I will be adding more vignettes when available!

***

## Interactive Maps in R with `leaflet`

I do a lot of mapping in R and have been using the `leaflet` package for a few years. Unlike static maps in `ggplot2`, these allow users to pan, zoom, and even click on map objects for more information. On one hand, the code required is a fairly gross mixture of R and JavaScript, but the package is extremely flexible and allows for a lot of customization. The following vignettes demonstrate some of the basic functionality of `leaflet` in R.

***

### *Basic Plotting*

```{r error = FALSE, message=FALSE, warning=FALSE, echo = FALSE}
{ list.of.packages <- c("tidyverse", "viridis", "knitr", "leaflet", "leafem", "leaflet.extras") # define necessary pkgs
  new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])] # compare necessary vs installed pkgs
    if(length(new.packages)) install.packages(new.packages) # install any new pkgs
  lapply(list.of.packages, library, character.only = TRUE) # load pkgs
  rm(list.of.packages, new.packages) # remove extra objects
  # sessionInfo() # check which packages are actually installed
  }

```

One of the things that's nice about `leaflet` is that it doesn't force you to format your dataset specifically for spatial datasets with `sp` or `sf`. You *can*, but it isn't required.

This first example was from a pilot study where we put GoPros down on a spawning reef to try getting footage of spawning Arctic Charr. We recorded camera number (*Camera*), GPS coordinates (*Latitude* & *Longitude*), and the direction the camera was facing (*Direction*). We also assigned a color to each camera based on whether it was on the reef or off the reef (*Color*).

The following code creates a simple interactive map with two layers: one for the GoPro cameras and one for some points of interest (POIs) that we identified in the area.

```{r error = FALSE, message=FALSE, warning=FALSE}
library(leaflet)

# Make a Test Dataset (Alternatively Import From File)
GoPro <- data.frame(Camera = as.character(c(11, 5, 7, 8, 4)),
                    Latitude =  c(44.73664, 44.73625, 44.73646, 44.73488, 44.73513),
                    Longitude = c(-68.48991,-68.48971,-68.48959,-68.49012,-68.49026),
                    Direction = c("West", "North", "West", "West", "South"),
                    Color = c("blue","blue","blue","green","green"))

GoPro_map <-
  leaflet(
    options = leafletOptions(zoomControl = FALSE) # turns off an annoying box
    ) %>% 
  # Add Basemap Layer (see https://leaflet-extras.github.io/leaflet-providers/preview/)
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") %>%
  
  # Add Points using GoPro data.frame 
  addCircleMarkers(
    # Define Dataset
    data = GoPro,
      lng = ~Longitude, # reference the coordinates columns
      lat = ~Latitude,  # reference the coordinates columns

    # Point Aesthetics
    color = ~Color,  # use column to reference color (could also make color palette function)
    stroke = FALSE,  # remove border for a cleaner look (at least I think so)
    radius = 6,      # adjust point size (can also adjust this based on a value)
    fillOpacity = 1, # make point opaque not transparent
    
    # Point Labeling
    popup = ~paste0("Direction: ", Direction, "<br>", # text displayed when clicking point
                    Latitude, ", ", Longitude),
    label = ~paste0("Direction: ", Direction),        # text displayed when hovering over point 
    group = "GoPros") %>%                             # layer group (important later)

  
  # Alternatively Add Points Manually
  addCircleMarkers( 
    data= data.frame(
      lng = c(-68.49053,-68.49260,
              -68.49139,-68.49579,-68.50065),
      lat = c( 44.73494, 44.73503,
               44.73659, 44.73776, 44.74542),
      label = c("Reef", "Potential Staging Area",
                "2024 Hotspot", "2024 Hotspot", "2024 Hotspot")
      ),
    lng = ~lng, # reference the coordinates columns
    lat = ~lat, # reference the coordinates columns
    
    color = "red",     # make all points in group a single color
    stroke = FALSE,    # remove border for a cleaner look
    fillOpacity = 0.5, # make points semi-transparent for fun
    
    popup = ~paste0(label, "<br>", lat, ", ", lng),
    label = ~label,
    group = "Points of Interest"
    ) 

GoPro_map

```
Play around with the map using your mouse and you'll notice a few things:

* pan by left clicking and dragging the map
* scrolling the mouse wheel or using `+` and `-` will adjust the zoom
* double clicking causes it to zoom in
* zooming out far enough changes the base map slightly based on image resolution
  +  this could also be adding/removing labels for lakes/cities/roads/etc depending on the base map

***

### *Toggle Layers and Basemaps*

Something that's also really important about `leaflet` is that (just like `ggplot`) layers are added in the order you call them, with the exception that basemaps are always on the bottom. You may notice that the red points of interest are kinda overlapping the green GoPro points, and eventually the blue points if you zoom out enough. Sometimes it's nice to be able to turn off layers or toggle between them

Notice above that in the code above, each layer added to the map above has an argument for `group`, such as `group = GoPros` and `group = World Imagery`. By using `addLayersControl` we can add `overlaygroups` that turn on/off and `baseGroups` that you can cycle between. The layer most recently turned on will be plotted on top, regardless of order they were coded for the control box.


```{r error = FALSE, message=FALSE, warning=FALSE}
GoPro_map %>%
  # Add More Basemap Options
  addProviderTiles(providers$OpenStreetMap.Mapnik, group = "Street Map") %>%
  addProviderTiles(providers$Stadia.AlidadeSmooth, group = "Grayscale") %>%
  
  addLayersControl(overlayGroups = c("GoPros", "Points of Interest"), # toggle these layers on/off
                   baseGroups = c("World Imagery", "Street Map", "Grayscale"), # switch between these
                   options = layersControlOptions(collapsed = FALSE))
```
I tend to put basemap layers as `baseGroups` (this is probably intended), but theoretically you could assign points or combinations of both. For example, add black/white points in the group "*Greyscale*" to have a potentially publication-friendly figure.

Look for [more basemap layers](https://leaflet-extras.github.io/leaflet-providers/preview/) to find one that suits your needs. Unfortunately, many require an API, so some trial and error may be necessary. I typically use layers from *OpenStreetMap*, *Stadia*, *Esri*, and *USGS*. Also be warned that sometimes layers will not zoom in as far.

*** 

### *Including Additional Tools*

Sometimes you might want to add some tools to make life easier for checking distances between points or simply drawing another (temporary) point or polygon. Here's how to add further functionality to maps using `leafem` and `leaflet.extras`

```{r error = FALSE, message=FALSE, warning=FALSE, eval = FALSE}
library(leafem)         # addMouseCoordinates()
library(leaflet.extras) # addDrawToolbar(), addMeasurePathToolbar(), addStyleEditor()

GoPro_map %>%

  # Display Lat & Long of Mouse Cursor and Zoom Level (0 is world view; these basemaps max out ~17-20)
  addMouseCoordinates() %>%

  # Simple Measuring Tool With a Lot of Customization (defaults to top right)
  addMeasure() %>%  # shows lat/long location (point), length (polyline), and area/perimeter (polygon)
                    # change length units: primaryLengthUnit/secondaryLengthUnit = "feet"/"miles
                    #                                                              "meters"/"kilometers"
                    # change area units: primaryAreaUnit/secondaryAreaUnit = "acres"/"hectares"
                    #                                                        "sqmeters"/"sqmiles"
                    # admittedly the default color schemes are awful
                      # change with completedColor/activeColor = "#000000" (use hex colors notation)

  
  # A More Complicated Drawing Toolset (the argument inside lets you edit vertices and delete objects)
  addDrawToolbar(editOptions=editToolbarOptions(selectedPathOptions=selectedPathOptions())) %>%
              # Symbols are for polylines, polygons, squares, circles, markers, and points  

    
  # Display Measurements of All Drawn Features (works for both drawing toolsets; defaults to left side)
  enableMeasurePath() %>%     # allows the display to work (default is meters)
  addMeasurePathToolbar() %>% # toggle on/off with beaker symbol; refresh after eding vertices
  

  # Manually Adjust Any Map Feature (plotted data or drawn line) 
  addStyleEditor() %>%  # Allows changes in point/line/polygons: 
                            # outline - color/opacity/thickness/dashes,
                            # fill - color/opacity, and label text (if clicked)
                        # this breaks markers (blue pins)

  
  # Add Search Bar for Specific Text in Popups (doesn't work on draw shapes/points)
  addSearchFeatures(targetGroups = c("GoPros", "Points of Interest")) # must define groups to search

```

```{r error = FALSE, message=FALSE, warning=FALSE, echo = FALSE}


GoPro_map %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik, group = "Street Map") %>%
  addProviderTiles(providers$Stadia.AlidadeSmooth, group = "Grayscale") %>%

  addLayersControl(overlayGroups = c("GoPros", "Points of Interest"),
                   baseGroups = c("World Imagery", "Street Map", "Grayscale"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  addMouseCoordinates() %>% 
  
  addMeasure() %>% 
  
  addDrawToolbar(editOptions=editToolbarOptions(selectedPathOptions=selectedPathOptions())) %>% # 

  enableMeasurePath() %>%
  addMeasurePathToolbar() %>%
  
    
  addStyleEditor() %>% # 
  
  addSearchFeatures(targetGroups = c("GoPros", "Points of Interest")) ##
  

```

Play around with some of these new tools! They typically default to the left side of the map (except `addMeasure()`), and are added top to bottom in order you add them in the code.The location of the toolbars for `addLayersControl()`, `addMeasure()`, `addDrawToolbar()` and `addStyleEditor()` can be adjusted using arguments like `position = "topleft"` or `position = "bottomright"`, though I have yet to figure out the syntax for `addMeasurePathToolbar()` and `addSearchFeatures()`





***

## Some of My Favorite R Code


***

### *Simultaneously Install & Load Packages*

Sometimes when sending code to collaborators (or myself on a different computer), it's important to make sure that all the packages you use are loaded. Though RStudio is pretty good about letting you know which aren't loaded, it's still easier to just write code to install and load packages rather than relying on the recipient to notice they don't have something!

This code checks for the packages you need from the list you provide, installs any that are missing, and loads them all to the work space. The packages used in this example were for the `leaflet` demos above.

I use variations of this code snippet in nearly every R script I write.

*Original*
```{r error = FALSE, message=FALSE, warning=FALSE, eval = FALSE}
install.packages("tidyverse")
install.packages("viridis")
install.packages("knitr")
install.packages("leaflet")
install.packages("leafem")
install.packages("leaflet.extras")

library(tidyverse)
library(viridis)
library(knitr)
library(leaflet)
library(leafem)
library(leaflet.extras)

```


*New*
```{r error = FALSE, message=FALSE, warning=FALSE, eval = FALSE}

{ # define necessary packages
  list.of.packages <- c("tidyverse", "viridis", "knitr", "leaflet", "leafem", "leaflet.extras")
  # compare necessary vs installed packages
  new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])] 
    if(length(new.packages)) install.packages(new.packages) # install any new pkgs
  lapply(list.of.packages, library, character.only = TRUE) # load pkgs
  rm(list.of.packages, new.packages) # remove extra objects
  # sessionInfo() # option to check which packages are actually installed
  }

```


***

### *Calculating Function Runtime*

I often run code that takes a long time to complete, so I wrote this function to calculate the runtime of any code you put inside it. Honestly it's a bit over-engineered, but this wrapper function was really helpful when I ran models that took anywhere from 20 minutes to three days to complete. Just wrap your code in the `runtime()` function and it will return a message with the total time elapsed.

```{r error = FALSE, message=FALSE, warning=FALSE}
runtime <- function(x){
  start <- Sys.time()
  x
  ifelse(difftime(Sys.time(), start) <  as.difftime(minutes(1)),
         paste0("Function ran for ", round(difftime(Sys.time(),start, units = "secs"),3),
                " seconds; Ending ", format(Sys.time(), format = "%B %d, %Y %H:%M:%S")),
         ifelse(difftime(Sys.time(),start) <  as.difftime(hours(2)),
                paste0("Function ran for ", round(difftime(Sys.time(),start, units = "mins"),2),
                       " minutes; Ending ", format(Sys.time(), format = "%B %d, %Y %H:%M:%S")),
                paste0("Function ran for ", round(difftime(Sys.time(),start, units = "hours"),2),
                       " hours; Ending ", format(Sys.time(), format = "%B %d, %Y %H:%M:%S"))
         ))
}

```

For example:
```{r error = FALSE, message=FALSE, warning=FALSE, echo = FALSE}
runtime(
  {
  5+3 
  Sys.sleep(2) # adds two seconds
  } 
)

```

***

### *Customizing `GGally::ggpairs()`*

I recently had to make a large set of pairwise plots of environmental covariates for a manuscript and ran into the problem of having too many plots to easily view them all. Secondly, I wanted to be able to quickly identify which plots had significant correlations.

I wrote this function in two parts: the first part updates `GGally::ggpairs` plots with customized lower panels to include correlation coefficients, p-values, is color coded. The second part allows you to subset these plots so a reasonable number can be viewed at a time.

```{r error = FALSE, message=FALSE, warning=FALSE}
library(GGally) # Yes, I realize that I'm not loading these using my fancy code above 
library(ggpubr)


ggpairs_updated <- function(covars){
  
  GGally::ggpairs(covars, lower = list(continuous = function(data, mapping, method="lm", ...){
    p <- ggplot(data = data, mapping = mapping) + 
      geom_point() + 
      geom_smooth(method=method,colour="blue", ...)+
      ggpubr::stat_cor(aes(label=paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
               method="pearson", label.y.npc = "center")
    
    
    x <- eval_data_col(data, mapping$x)
    y <- eval_data_col(data, mapping$y)
    
    # Calculate correlation and p-value
    test_res <- cor.test(x, y, use = "complete.obs")
    p_val <- test_res$p.value
    r_val <- test_res$estimate
    
    # Determine color based on p-value (e.g., p < 0.01 is green, p < 0.05 is yellow, otherwise grey)
    # You can customize these colors and thresholds
    if (p_val < 0.01) {
      border_col <- "green3"
      border_lw <- 3
      bg_col <- "green3"
    } else if (p_val < 0.05) {
      border_col <- "gold"
      border_lw <- 2
      bg_col <- "gold"
    } else {
      border_col <- "grey92"
      border_lw <- 1.5
      bg_col <- "grey92"
    }
    
    p <- p + theme(#panel.border = element_rect(fill = NA, color = border_col, linewidth = border_lw),
      panel.background = element_rect(fill = bg_col, color = NA))
    
    return(p)
    
  }
  ))
}

```

Example using the entire `mtcars` dataset
```{r error = FALSE, message=FALSE, warning=FALSE, echo = FALSE}
library(cowplot)

plot_grid(
  ggmatrix_gtable(ggpairs(mtcars)),
  ggmatrix_gtable(ggpairs_updated(mtcars)),
  ncol = 2,
  labels = c("Old ggpairs", "Updated ggpairs"))

```

Again, this second part takes the output from `ggpairs_updated()` and allows you to subset the plots by row and column. This is useful when you have a large number of variables and want to view only a few at a time.

```{r error = FALSE, message=FALSE, warning=FALSE}
ggpairs_subset <- function (ggpairs_output, selected_rows = NULL, selected_cols = NULL){
  
  plot_list = list()
  
  if (is.null(selected_rows)){selected_rows <- 1:ggpairs_output$nrow}
  if (is.null(selected_cols)){selected_cols <- 1:ggpairs_output$ncol}
  
  for (i in seq_along(selected_rows)) {
    for (j in seq_along(selected_cols)) {
      # Extract the specific panel using getPlot
      p <- getPlot(ggpairs_output, i = selected_rows[i], j = selected_cols[j])
      
      # Store in list
      plot_list <- append(plot_list, list(p))
    }
  }
  
  output <- ggmatrix(
    plots = plot_list,
    nrow = length(selected_rows),
    ncol = length(selected_cols),
    xAxisLabels = ggpairs_output$xAxisLabels[selected_cols],
    yAxisLabels = ggpairs_output$yAxisLabels[selected_rows]
  )
  
  return(output)
  # rm(plot_list, selected_rows, selected_cols, p, plot_list, i, j, output)
}
```

Subset `mtcars` to only show specific rows & columns; these don't need to be the same rows and columns
```{r error = FALSE, message=FALSE, warning=FALSE, echo = FALSE}
ggpairs_subset(ggpairs_updated(mtcars), selected_rows = c(1,3:7), selected_cols = c(1,3:7))

```



***

## Future topics

Some potential additions include:

- customizing `leaflet` point markers using *Awesome Markers*
- estimating survival and other demographic parameters in `RMark`
- interpolating lake bathymetry from points
- calculating percent lake volume below/above depth contours 
